---
title: "Assignment_4"
author: "Tarun BT"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
Imported necessary library files required for task
```{r}
library(ggplot2)
library(lattice)
library(caret)
library(e1071)
library(dplyr)
library(pROC)
library(naivebayes)

```

Loading the data set
```{r}
Heart_disease <- read.csv("C:\\Users\\Tarun\\OneDrive\\Desktop\\R program\\4th Assignment\\Heart_disease.csv")

```

Checking for duplicate value and removing record from the data-set
```{r}
duplicates <- Heart_disease[duplicated(Heart_disease), ]
cat("Number of duplicate record ",nrow(duplicates),"\n")
print(duplicates) 

df_unique <- Heart_disease %>% distinct() #data-set after removing the duplicate value
nrow(df_unique)

```

Creating dummy variables
```{r}
df_unique$Target <- ifelse(df_unique$MAX_HeartRate > 170, "Yes", "No")
df_unique$BP_New <- ifelse(df_unique$Blood_Pressure > 120, "Yes", "No")

```

Q1: Initial Prediction Based on Target Distribution
```{r}
target_table <- table(df_unique$Target)
target_table

```

Bar Plot for Target Distribution
```{r}
ggplot(df_unique, aes(x = Target)) + geom_bar(fill = "steelblue") + ggtitle("Distribution of Target Variable")

```
Interpretation:
No appeared 245 counts
Yes appeared 57 counts
Since No is the majority class - No Heart Disease


Q2: Analysis of the First 30 Records
```{r}
Heart_disease30 <- df_unique[1:30, c("Target", "BP_New", "chest_pain_type")]
Object1 <- ftable(Heart_disease30)
Object1
head(Heart_disease30)

```
pivot table without target column
```{r}
Object2 <- ftable(Heart_disease30[ ,-1])
Object2 

```

2a. Computing Bayes Conditional Probabilities
```{r}
p1 <- Object1[3,1]/Object2[1,1] #Target=yes , BP_New=No & chest_pain_type=0
p2 <- Object1[3,2]/Object2[1,2] #Target=yes , BP_New=No & Chest_pain_type=1
p3 <- Object1[4,1]/Object2[2,1] #Target=yes , BP_New=Yes & Chest_pain_type=0
p4 <- Object1[4,2]/Object2[2,2] #Target=yes , BP_New=Yes & Chest_pain_type=1

```

Conditional probabilities values
```{r}
cat("The conditional probability of having heart disease with no BP and No chest Pain:",p1,"\n")
cat("The conditional probability of having heart disease with high BP and chest Pain:", p2,"\n")
cat("The conditional probability of having heart disease with BP and No chest Pain:", p3 ,"\n")
cat("The conditional probability of having heart disease with BP and chest Pain:", p4 ,"\n")

```


2b. Classification using a cutoff of 0.5
```{r}
Probability_Target <- rep(0, 30)

for (i in 1:30) {
  BP_New <- Heart_disease30$BP_New[i]  # Getting BP_New value for row i
  chest_pain <- Heart_disease30$chest_pain_type[i]  # Getting chest_pain for row i
  
  # Matching the probability from pre-computed values (p1, p2, p3, p4)
  if (BP_New == "No" & chest_pain == 0) {
    Probability_Target[i] <- p1
  } else if (BP_New == "Yes" & chest_pain == 1) {
    Probability_Target[i] <- p2
  } else if (BP_New == "Yes" & chest_pain == 0) {
    Probability_Target[i] <- p3
  } else if (BP_New == "No" & chest_pain == 1) {
    Probability_Target[i] <- p4
  }
}

Heart_disease30$Probability_Target <- Probability_Target
Heart_disease30$Pred_Probability <- ifelse(Heart_disease30$Probability_Target > 0.5, "Yes", "No")
Heart_disease30
```

2c. Manual Calculation of Naive Bayes Probability
```{r}
# Compute total instances
total_count <- nrow(Heart_disease30)

# Compute P(Target = Yes)
p_target_yes <- sum(Heart_disease30$Target == "Yes") / total_count

# Compute P(BP_New = Yes, chest_pain_type = 1 | Target = Yes)
p_given_yes <- sum(Heart_disease30$BP_New == "Yes" & Heart_disease30$chest_pain_type == 1 & Heart_disease30$Target == "Yes") / sum(Heart_disease30$Target == "Yes")

# Compute P(BP_New = Yes, chest_pain_type = 1)
p_denominator <- sum(Heart_disease30$BP_New == "Yes" & Heart_disease30$chest_pain_type == 1) / total_count

# Compute the final probability using Bayes' Theorem
p_yes_given_bp_chest <- (p_given_yes * p_target_yes) / p_denominator

# Print result
cat("Manually compute the naive  Bayes conditional probability of an injury given that BP_New is Yes and chest_pain_type is 1 = ",p_yes_given_bp_chest, "\n")

```
Q3: Full Data set Analysis - Splitting into Training and Validation Sets
```{r}
set.seed(123)

train.index <- sample(row.names(df_unique),0.6 * nrow(df_unique))
valid.index <- setdiff(row.names(df_unique),train.index)

train.df <- df_unique[train.index, ]
valid.df <- df_unique[valid.index, ]

nrow(train.df)
nrow(valid.df)

train.df <- train.df[,-9]
valid.df <- valid.df[,-9]

```

Running Naive Bayes Classifier 
```{r}
nb_model <- naiveBayes(Target ~ chest_pain_type + BP_New, data = train.df,1)
valid_pred <- predict(nb_model, valid.df)

# Convert predictions and actual target variable to factors with the same levels
valid.df$Target <- factor(valid.df$Target)
valid_pred <- factor(valid_pred, levels = levels(valid.df$Target))

# Compute confusion matrix
conf_matrix1 <- confusionMatrix(valid_pred, valid.df$Target)

```

```{r}
conf_matrix1
```

Running Naive Bayes Classifier including all the attributes. (Validation set)
```{r}
nb_model_1 <- naiveBayes(Target ~., data = train.df)
train_pred_1 <- predict(nb_model_1, valid.df)

# Convert predictions and actual target variable to factors with the same levels
valid.df$Target <- factor(valid.df$Target)
train_pred_1 <- factor(train_pred_1, levels = levels(valid.df$Target))

# Compute confusion matrix
conf_matrix1_1 <- confusionMatrix(train_pred_1, valid.df$Target)
conf_matrix1_1
```

ROC Curve
```{r}
pred_probs_full <- predict(nb_model_1, valid.df, type = "raw")
prob_yes_full <- pred_probs_full[, "Yes"]
roc_full <- roc(valid.df$Target, prob_yes_full)
plot(roc_full, col = "blue", main = "ROC Curve for All Attributes")
auc_value <- auc(roc_full)
auc_value
```