---
title: "1st_ML"
author: "Tarun BT"
date: "`r Sys.Date()`"
output: html_document
---
## 1. Conduct basic descriptive statistics, including head, tail, dimensions, summary, structure, mean, mode, median, and check for any missing values.
```{r setup, include=TRUE}
library(dplyr)
library(readr)
library(psych)
library(modeest)
library(tidyverse)
library(ggplot2)
library(tinytex)

data <- read_csv("C:\\Users\\Tarun\\OneDrive\\Desktop\\R program\\Churn.csv")
dim(data) #to find the dimension of data
summary(data) #to find the summary of data
str(data) #to find the structure of data
describe(data) 

head(data) #to display the head of dataset
tail(data) #to display the tail of dataset

numeric_cols <- data %>% select_if(is.numeric) #to extract the numeric value columns 

mean_values <- sapply(numeric_cols, mean, na.rm = TRUE) #na.rm=TRUE to ignore the missing values 
median_values <- sapply(numeric_cols, median, na.rm = TRUE) 
mode_values <- sapply(numeric_cols, function(y) mfv(y, na_rm = TRUE)) #function(y)-Applies a function to each column of the data dataframe. mfv - most frequent value

which.max(table(data$ContractType)) #Finds the most frequent ContractType category in the dataset

missing_values <- sapply(data, function(y) sum(is.na(y))) #Counts missing (NA) values in each column of the dataset

list(
  Mean = mean_values,
  Median = median_values,
  Mode = mode_values,
  Missing = missing_values
)  # Creates a list storing mean, median, mode, and missing value counts for each column
```

## Questio-2 
The data set used here contains nine columns and 1000 rows. The first intention of the current analysis is to examine the impact of various factors on customer attrition and mine knowledge with the help of descriptive statistics. The examination of data structure identified that apart from categorical Type-"Contract" column, all the variables are numeric. Then the data set was checked using the is.na command to determine if there were any missing values. They were discovered to be none. In this case, descriptive statistics were utilized to determine additional information about the behavior of the customers. Summary statistics of the numerous variables in the data set, such as Mean, Median, and Mode, are given in the summary table. As the mean and median values for most of the variables are similar, the data is well balanced and does not contain many discrepancies. The mean value of churn is 0.504, i.e., in between 50 and 50 percent of the customers are remaining or leaving. Clients can be leaving due to the cost if the average monthly charge for the clients who have left is greater. Data Use, Voice Calling Minutes, Charges per Month, and Roaming Use did not have a mode when we studied the mode for each variable since all values were distinct. Churn mode was 1 and indicated that most customers were exiting the business as opposed to retaining it. Most frequently used type of contract being selected by the customers was monthly type, even according to mode for contract type. Additionally, the fact that the majority of customers call customer support more than twice points to ongoing or unresolved issues. Extreme differences between minimum and maximum data usage must be investigated.

## 3.Select and compare two or more variables from the dataset, such as DataUsage and Churn. Create a table for the  selected variables using the df(dataframe) and select commands.

```{r pressure, echo=TRUE}
# Provide summary statistics or visualizations to support your comparison. 

selected_data <- data %>% select(DataUsage, Churn) #selecting the variables from the csv file
view(selected_data) #to view the selected data in the table format
print(selected_data) # to print the table details.

cor_value <- cor(selected_data$DataUsage,selected_data$Churn) #Calculates the correlation between Data Usage and Churn to measure their relationship strength and direction.

print(paste("Co-relation value is", cor_value))

ggplot(selected_data, aes(x = as.factor(Churn), y = DataUsage, fill = as.factor(Churn))) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  labs(x = "Churn (0 = No, 1 = Yes)", y = "Data Usage",
       fill = "Churn", title = "Boxplot of Data Usage by Churn") +
theme_minimal() # Creates a boxplot to visualize the distribution of Data Usage for each Churn category (0 = No, 1 = Yes), highlighting outliers in red.
```

Conclusion:
The analysis suggests no strong relationship between DataUsage and Churn.
Both boxplots and correlation values confirm this
Other factors like contract type, customer service calls, or pricing might be better indicators of churn

## 4. Transform at least one variable (any transformation such  as log(x) and generate a plot using ‘ggplot’ for the transformed variable.
```{r, echo=TRUE}

data <- data %>% mutate(LogDataUsage = log(DataUsage + 1)) # Creates a new column 'LogDataUsage' by applying a log transformation to 'DataUsage' to normalize its distribution.


ggplot(data, aes(x = LogDataUsage)) +
  geom_density(fill = "lightblue", color = "darkblue", alpha = 0.7) +
  labs(x = "Log(Data Usage)", y = "Density", 
       title = "Density Plot of Log-Transformed Data Usage") +
  theme_minimal() # Generates a density plot to visualize the distribution of log-transformed Data Usage, with a light blue fill and dark blue border.
```
