---
title: "2nd_Assignment"
author: "Tarun BT_811390248"
date: "`r Sys.Date()`"
output: html_document
---

## 2. Assignment 
Based on “CARET Package (5:03 minutes)” video in Module 3, modify your code for the Carseats example. We would like to have three sets of data, Training (~200 records), Validation (~100 records) and Test (~100 records). Make sure any record/row of data appears only once in either of these three classes (no duplication).

```{r,}
library(ggplot2)
library(lattice) #imported the libraries to avoid the warring message.
```

```{r setup, include=T}
library(ISLR) # Contains the Several Data sets (Carseats Data set)
library(caret) #Package designed for machine learning workflows and provides functions for data preprocessing, feature selection, model training, and evaluation.
```

```{r,}
data("Carseats") #load the Carseats dataset
```

# hint 1: 
First,  partition Carseats dataset into Temp (about 300 records) and Test (100 records).
```{r,}
trainIndex <- createDataPartition(1:nrow(Carseats), p = 0.75, list = FALSE)  # 75% for Temp
Temp <- Carseats[trainIndex, ]  # 'Temporary' set containing 300 records
Test <- Carseats[-trainIndex, ] # 'Test' set containing 100 records 
```

# hint 2:
Then, partition Temp into two distinct datasets, i.e., Training (200 records) and Validation (100 records).
```{r}
trainIndex2 <- createDataPartition(1:nrow(Temp), p = 2/3, list = FALSE)  # p=2/3 is assigned for Training
Training <- Temp[trainIndex2, ]  # 'Training' set containing 200 records
Validation <- Temp[-trainIndex2, ] # 'Validation' set containing 100 records
```

# Display summary statistics for each dataset
```{r}
summary(Training)
summary(Validation)
summary(Test)
```

# Check the dimensions of each dataset
```{r}
cat("Training set dimensions(Rows,Columns):", dim(Training), "\n") # cat() function is used format the output in the structured way.
cat("Validation set dimensions(Rows,Columns):", dim(Validation), "\n")
cat("Test set dimensions(Rows,Columns):", dim(Test), "\n")
```
# To view the dataset.
```{r}
View(Training)
View(Validation)
View(Test)
```

```{r}
duplicates_all_three <- intersect(intersect(Training, Validation), Test)

print(duplicates_all_three)
```

