---
title: "1st_ML"
author: "Tarun BT_811390248"
date: "`02/03/2025`"
output:
  html_document: default
  pdf_document: default
---
## 1.Conduct basic descriptive statistics, including head, tail, dimensions, summary, structure, mean, mode, median, and check for any missing values.
```{r setup, include=TRUE}
library(dplyr)
library(readr)
library(psych)
library(modeest)
library(tidyverse)
library(ggplot2)
library(tinytex)
```

```{r,echo=T}
#reading file
data <- read_csv("C:\\Users\\Tarun\\OneDrive\\Desktop\\R program\\Churn.csv")
```

```{r,echo=TRUE}
dim(data) #to find the dimension of data
```

```{r,echo=T}
summary(data) #to find the summary of data
```

```{r,echo=TRUE}
str(data)#to find the structure of data
```

```{r,echo=TRUE}
describe(data) 
```

```{r,echo=TRUE}
head(data) #to display the head of dataset
```

```{r,echo=TRUE}
tail(data)  #to display the tail of datase
```

```{r,echo=TRUE}
numeric_cols <- data %>% select_if(is.numeric) #to extract the numeric value columns
```

```{r,echo=TRUE}
mean_values <- sapply(numeric_cols, mean, na.rm = TRUE) #na.rm=TRUE to ignore the missing values 
print(mean_values)
```

```{r,echo=TRUE,}
median_values <- sapply(numeric_cols, median, na.rm = TRUE) 
print(median_values)
```

```{r,echo=TRUE}
mode_values <- sapply(numeric_cols, function(y) mfv(y, na_rm = TRUE)) #function(y)-Applies a function to each column of the data dataframe. mfv - most frequent value
print(mode_values)
```

```{r,echo=TRUE}
which.max(table(data$ContractType)) #Finds the most frequent ContractType category in the dataset
```

```{r,echo=TRUE}
missing_values <- sapply(data, function(y) sum(is.na(y))) #Counts missing (NA) values in each column of the dataset
print(missing_values)
```

##  2.Write an interpretation of your analysis in part 1(descriptive statistics) in R Markdown report.

The data set used here is 1000 rows and 9 columns, or 1000 observations and 9 variables in R terms. The main goal of this analysis is to investigate how different factors affect customer attrition and extract knowledge using descriptive statistics. Analysis of the data structure revealed that, except for the Contract Type column, which was categorical, all of the variables were numerical. Key statistics for the many variables in the dataset, including the mean, median, and quartiles, are displayed in the summary table. Here, descriptive statistics were employed to learn more about the behavior of the customers. Key statistics, including mean, median, and quartiles for each variable in the dataset, are displayed in the summary table. Descriptive statistics were used to learn more about the behavior of the customers. The is.na command was used to check the dataset for missing values, and the results showed that there were none. The mean value for monthly charges is 58.41, indicating that a fair number of customers are paying higher amounts. With a mean churn value of 0.504, almost half of the customers remain while the other half depart. Additionally, the churn mode was 1, indicating that more consumers were leaving the business than were remaining. Customers experience up to 10 call failures on average, according to the mean number for call failures, which is 10. Some consumers have never experienced a call failure, but others have, and this can lead to churn and customer discontent. Data usage, voice minutes, monthly charges, and roaming usage were found to have no mode when the mode for each variable was examined because each value was distinct.The most popular contract type selected by clients was the monthly type, according to the mode for contract type as well. Additionally, it was noted that while the most frequent number of customer care calls was zero, on average, clients had contacted customer assistance at least twice, indicating persistent or unresolved concerns. There is a concerning discrepancy between the minimum and maximum data usage, which calls for more research.

## 3.Select and compare two or more variables from the dataset, such as DataUsage and Churn. Create a table for the  selected variables using the df(dataframe) and select commands.

```{r pressure, echo=TRUE}
# Provide summary statistics or visualizations to support your comparison. 

selected_data <- data %>% select(DataUsage, Churn) #selecting the variables from the csv file
view(selected_data) #to view the selected data in the table format
print(selected_data) # to print the table details.

cor_value <- cor(selected_data$DataUsage,selected_data$Churn) #Calculates the correlation between Data Usage and Churn to measure their relationship strength and direction.

print(paste("Co-relation value is", cor_value))

ggplot(selected_data, aes(x = as.factor(Churn), y = DataUsage, fill = as.factor(Churn))) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 2) +
  labs(x = "Churn (0 = No, 1 = Yes)", y = "Data Usage",
       fill = "Churn", title = "Boxplot of Data Usage by Churn") +
theme_minimal() # Creates a boxplot to visualize the distribution of Data Usage for each Churn category (0 = No, 1 = Yes), highlighting outliers in red.
```

Conclusion:
Churn and DataUsage do not appear to be strongly correlated, according to the analysis.
This is supported by correlation values and boxplots.
Other elements such as pricing, customer service calls, CallFailures, SubscriptionLength, VoiceMinutes CustomerSupportCalls or contract type may be more accurate churn indicators.


## 4.Transform at least one variable (any transformation such  as log(x) and generate a plot using ‘ggplot’ for the transformed variable.
```{r, echo=TRUE}

data <- data %>% mutate(LogDataUsage = log(DataUsage + 1)) # Creates a new column 'LogDataUsage' by applying a log transformation to 'DataUsage' to normalize its distribution.


ggplot(data, aes(x = LogDataUsage)) +
  geom_density(fill = "lightblue", color = "darkblue", alpha = 0.7) +
  labs(x = "Log(Data Usage)", y = "Density", 
       title = "Density Plot of Log-Transformed Data Usage") +
  theme_minimal() # Generates a density plot to visualize the distribution of log-transformed Data Usage, with a light blue fill and dark blue border.
```
